Characterizing dynamical transitions by statistical complexity measures based on ordinal pattern transition networks

### 2020-10-28 manuscript_Huang.tex
updated.

\textcolor{red}{If this paper is dedicated for submission to Chaos, a lead paragraph needs to be added here.}

\section{Introduction}
In the recent decade, complex network based approaches have emerged as prominent tools for nonlinear time series analysis \cite{Kantz97,Sprott2003}, which have found many interesting applications to experimental data \cite{ZouPR2018}. For instance,

There are several major methods are of particular importance, including recurrence network \cite{MarwanPLA2009,Donner2010a}, visibility graphs \cite{Lacasa2008,Nunez2012}, and transition networks \cite{Nicolis2005} and cycle network \cite{ZhangPRL2006}. Depending on the particular questions at hand, we note that each of these complex network approaches has various variants for different applications, for instance, constructing a single network for a univariate time series or interacting networks (multiplex or multilayer) networks for coupled time series. This line of research is undergoing fast developments for the time being as reviewed in \cite{ZouPR2018}.

The focus of this paper is ordinal pattern transition networks. The advantages and some recent results.

@@

Note that, in all above discussions of ordinal pattern analysis, the transition behavior between patterns remains largely unclear. Recently, ordinal partition network representations have been proposed to capture the transition behavior of the ordinal patterns \cite{MichaelChaos2015,KulpChaos2016,zhangSciRep2017,McCullough2017b,BorgesAMC2019}, which sheds novel insight on the standard ordinal symbolic analysis of time series. In particular, each ordinal pattern is considered as a vertex in the graph and a directed weighted edge connecting two patterns in the graph is established depending on the transition frequency. One of advantages of ordinal pattern transition networks is that we can provide pronounced distinction among different time series with short time series \cite{MichaelChaos2015,BorgesAMC2019}.

@@

\section{Choice of embedding parameters} \label{sec:embeddings}
There are two important algorithmic parameters in constructing ordinal pattern transition networks, namely, embedding parameters of $D$ and $\tau$. In some cases, the choice of $D$ and $\tau$ may strongly change the results since $D$ and $\tau$ should compromise the time scales of the respective time series. In general applications of ordinal pattern analysis (i.e., permutation entropy computation $\mathcal{S}_O$), the choice of optimal $D$ and $\tau$ depends on the field of application, which requires experience of experts as systematically summarized in \cite{Riedl2013}.
\begin{figure*}
	\centering 
	\includegraphics[width=2\columnwidth]{period4_logisticExample.pdf}
\caption{\small{Example of a period-4 iterative time series ($r= 3.5$ in the Logistic map) to show the effects of embedding parameters $D$ and $\tau$. (a) iterative time series and the corresponding ordinal pattern series by different $D$ and $\tau$, (b) $D = \tau = 3$, (c) $D = \tau = 4$, (d) $D = \tau = 5$, and (e) $D = \tau = 6$. The respective permutation entropy values normalized by $\log D!$ as Eq. \eqref{eq:Ho} are $\mathcal{H}_O = 0.77$, $0$, $0.29$, and $0.11$ as indicated in the insets of each panel. (f-i) show the corresponding ordinal pattern transition networks. The weights of directed edges are suppressed due to two slightly different normalizations are introduced. Note in (g), only self-loops are available since $D = \tau = 4$ result in a unique pattern. Missing patterns are suppressed. }
\label{fig:embed}}
\end{figure*}
Here we use one example to show the important roles of $D$ and $\tau$ for a period-4 time series in Figure \ref{fig:embed}. In this example, we choose $D = \tau $ and the corresponding ordinal pattern series are significantly influenced by these parameter settings, which further result in different permutation entropy values of $\mathcal{H}_O$.  For instance, only one unique ordinal pattern is observed when $D = \tau = 4$ (Fig. \ref{fig:embed}(c, g)) resulting in $\mathcal{H}_O = 0$, which is due to the coverage of one full period by the particular embedding parameters. Other values of $D$ and $\tau$ yield non-zero entropy values. The results of Fig. \ref{fig:embed} seem to be pessimistic since $D$ and $\tau$ will change the positions in the complexity-entropy planes. In the special case of the logistic map, we do not have a unique choice of $D$ and $\tau$ for periodic windows when the control parameter $r$ is varied. In such cases, we should focus more on the transitions between different dynamic regimes. Instead, we show that the results are robust when the control parameters are systematically changed while using the same parameter settings of $D$ and $\tau$.

@@

\section{Discussions} \label{sec:con}
complex network approaches for nonlinear time series analysis \cite{ZouPR2018}.

>>>>

\begin{quotation}
In the recent decade, tools of traditional nonlinear time series analysis have been undergoing fast developments benefiting concepts from complex network theories. Along this line of research topics, ordinal pattern transition time series network has been providing complement insights to the permutation entropy estimators for time series data. Traditionally, permutations based on order patterns are simple and easily implemented tools for estimators of statistical complexity measures (SCMs), which however rely on the static pattern frequencies only. Still, much additional knowledge of the order pattern transitions to the SCMs has not been fully considered. In this work, we generalize the estimators of SCMs by means of ordinal pattern transition networks which take into account the pattern transition properties explicitly. The usefulness of the generalizations will be demonstrated by time series of both model and experimental data. 
\end{quotation}

\section{Introduction}
In the recent decade, complex network based approaches have emerged as prominent tools for nonlinear time series analysis \cite{Kantz97,Sprott2003}, which have found many interesting applications to experimental data \cite{ZouPR2018}. In this line of research, there are several major methods are of particular importance, including recurrence network \cite{MarwanPLA2009,Donner2010a}, visibility graphs \cite{Lacasa2008,Nunez2012}, transition networks \cite{Nicolis2005,MichaelChaos2015} and cycle network \cite{ZhangPRL2006}. Depending on the particular questions at hand, we note that each of these complex network approaches has various variants for different applications, for instance, constructing a single network for a univariate time series or interacting networks (multiplex or multilayer) networks for coupled time series, which has been systematically reviewed in \cite{ZouPR2018} recently. In this work, we focus on constructing ordinal pattern transition networks for time time series data, which show the advantages of easy implementations and wide applications to various data from different origins, including correlated stochastic processes, EEG and human cardiac series (ECG) \cite{MichaelChaos2015,KulpChaos2016,zhangSciRep2017,McCullough2017b,BorgesAMC2019}.

@@

Note that, in all above discussions of ordinal pattern analysis, the transition behavior between patterns remains largely unconsidered. Recently, ordinal partition network representations have been proposed to capture the transition behavior of the ordinal patterns \cite{MichaelChaos2015,KulpChaos2016,zhangSciRep2017,McCullough2017b,BorgesAMC2019}, which sheds novel insight on the standard ordinal symbolic analysis of time series. In particular, each ordinal pattern is considered as a vertex in the graph and a directed weighted edge connecting two patterns in the graph is established depending on the transition frequency. One of advantages of ordinal pattern transition networks is that we can provide pronounced distinction among different time series with short time series \cite{MichaelChaos2015,BorgesAMC2019}.

@@

\section{Choice of embedding parameters} \label{sec:embeddings}
{\color{red}We may leave out this section, if we plan a second paper which deals with embedding issue explicitly. }
There are two important algorithmic parameters in constructing ordinal pattern transition networks, namely, embedding parameters of $D$ and $\tau$. In some cases, the choice of $D$ and $\tau$ may strongly change the results since $D$ and $\tau$ should compromise the time scales of the respective time series. In general applications of ordinal pattern analysis (i.e., permutation entropy computation $\mathcal{S}_O$), the choice of optimal $D$ and $\tau$ depends on the field of application, which requires experience of experts as systematically summarized in \cite{Riedl2013}.
\begin{figure*}
	\centering 
	\includegraphics[width=2\columnwidth]{period4_logisticExample.pdf}
\caption{\small{Example of a period-4 iterative time series ($r= 3.5$ in the Logistic map) to show the effects of embedding parameters $D$ and $\tau$. (a) iterative time series and the corresponding ordinal pattern series by different $D$ and $\tau$, (b) $D = \tau = 3$, (c) $D = \tau = 4$, (d) $D = \tau = 5$, and (e) $D = \tau = 6$. The respective permutation entropy values normalized by $\log D!$ as Eq. \eqref{eq:Ho} are $\mathcal{H}_O = 0.77$, $0$, $0.29$, and $0.11$ as indicated in the insets of each panel. (f-i) show the corresponding ordinal pattern transition networks. The weights of directed edges are suppressed due to two slightly different normalizations are introduced. Note in (g), only self-loops are available since $D = \tau = 4$ result in a unique pattern. Missing patterns are suppressed. }
\label{fig:embed}}
\end{figure*}
Here we use one example to show the important roles of $D$ and $\tau$ for a period-4 time series in Figure \ref{fig:embed}. In this example, we choose $D = \tau $ and the corresponding ordinal pattern series are significantly influenced by these parameter settings, which further result in different permutation entropy values of $\mathcal{H}_O$.  For instance, only one unique ordinal pattern is observed when $D = \tau = 4$ (Fig. \ref{fig:embed}(c, g)) resulting in $\mathcal{H}_O = 0$, which is due to the coverage of one full period by the particular embedding parameters. Other values of $D$ and $\tau$ yield non-zero entropy values. The results of Fig. \ref{fig:embed} seem to be pessimistic since $D$ and $\tau$ will change the positions in the complexity-entropy planes. In the special case of the logistic map, we do not have a unique choice of $D$ and $\tau$ for periodic windows when the control parameter $r$ is varied. In such cases, we should focus more on the transitions between different dynamic regimes. Instead, we show that the results are robust when the control parameters are systematically changed while using the same parameter settings of $D$ and $\tau$.

@@

\section{Discussions} \label{sec:con}
In summary, we have proposed to include ordinal pattern transitions into the traditional ordinal pattern frequency analysis, generalizing statistical complexity measures for nonlinear time series analysis. In terms of the pattern transitions in the time series networks, permutation entropy relies on the static pattern frequency only. Here we suggest two slightly different ways to consider the effects of pattern transitions which are based on (i) a globally normalized weight transition matrix $\mathbf{W}$ and (ii) node-wise normalized transition matrix.

complex network approaches for nonlinear time series analysis \cite{ZouPR2018}.
